\chapter Changing layout.

This chapter is extremely useful if you're not satisfied with the predefined 
layout of map symbols and maps provided, and want to adapt them to your needs.
However, you need to know how to write plain \TeX\ and \MP\ macros to do this.


\subchapter Page layout in the atlas mode.

The |layout| command allows basic page setup in the atlas mode. This is done 
through its options such as |page-setup| or |overlap|. But there are no 
options which would specify the position of map, navigator and other elements 
inside the area defined by |page-width| and |page-height| dimensions; e.g., 
why is the navigator below the map and not on its right or left side?

There are many possible arrangements for a page. Rather than offer even more
options for the |layout| command, Therion uses the \TeX\ language to 
describe other page layouts. 
%(See the section `layout' where to put \TeX\ commands.) 
This approach has the advantage that the user has 
direct access to the advanced typesetting engine without making the language of 
Therion overcomplex.

Therion uses pdf\TeX\ with the {\it plain} format for typesetting. So you 
should be familiar with the plain \TeX\ if you wish to define new layouts.

The ultimate reference for plain \TeX\ is
\list
  Knuth, D. E.: {\it The \TeX book}, 
                Reading, Massachusetts, Addison-Wesley $^1$1984 %[...]
\endlist

For pdf\TeX's extensions there is a short manual
\list
  Th\`anh, H. T.---Rahtz, S.---Hagen, H.: {\it The pdf\TeX\ user manual}, 
                available at \path|http://www.pdftex.org|
\endlist

The \TeX\ macros are used inside of |layout| command (see the chapter 
{\it Processing data} for details). The basic one predefined by Therion is the

|\dopage|

macro. The idea is simple: for each page Therion defines \TeX\ variables 
(count, token, and box registers) which contain the page elements (map, 
navigator, page name etc.). At the end of each page macro |\dopage| is invoked. 
This defines the position of each element on the page. By redefining this macro 
you'll get desired page layout. Without this redefinition you'll get the 
standard layout.

Here is the list of variables defined for each page:

\list
  {\it Boxes:}

  * |\mapbox| = The box containing the map. Its width (height) is set 
    according to the |size| and |overlap| options of the |layout| command to 

    |size_width + 2*overlap| or 
    
    |size_height + 2*overlap|, respectively

  * |\navbox| = The box containing the navigator, with dimensions

    |size_width * (2*nav_size_x+1) / nav_factor| or

    |size_height * (2*nav_size_y+1) / nav_factor|, respectively

    Both |\mapbox| and |\navbox| also contain hyperlinks.

  {\it Count registers:}

  * |\pointerE|, |\pointerW|, |\pointerN|, |\pointerS| contain the page number 
    of the neighbouring pages in the E, W, N and S directions. If there is no 
    such a page its page number is set to 0.

  * |\pagenum| current page number

  {\it Token registers:}

  * |\pointerU|, |\pointerD| contain information about pages above and below
    the current page. It consists of one or more concatenated records. Each 
    record has a special format
    
    {\tt|page-name|\char124|page-number|\char124|destination|\char124\char124}
    
    If there are no such pages, the value is set to |notdef|.

    See the description of the |\processpointeritem| macro below for how to 
    extract and use this information.

  * |\pagename| = name of the current map according to options of the |map|
    command.

  * |\pagelabel| = the page label as specified by |origin| and |origin-label|
    options of the |layout| command.
\endlist


The following variables are set at the beginning of the document:

\list
  * |\hsize|, |\vsize| = \TeX\ page dimensions, set according to |page-width|
    and |page-height| parameters of the |page-setup| option of the |layout|
    command. They determine our playground when defining page layout using 
    the |\dopage| macro.

  * |\ifpagenumbering| = This conditional is set true or false according to
    the |page-numbers| option of the |layout| command.
\endlist

There are also some predefined macros which help with the processing of 
|\pointer*| variables:

\list
  * |\showpointer| with one of the |\pointerE|, |\pointerW|, |\pointerN| or 
    |\pointerS| as an argument displays the value of the argument. If the value
    is 0 it doesn't display anything. This is useful because
    the zero value (no neighbouring page) shouldn't be displayed.
    
  * |\showpointerlist| with one of the |\pointerU| or |\pointerD| as an argument 
    presents the content of this argument. (Which contains 
    |\pointerU| or |\pointerD|, see above.) For each record it calls the macro 
    |\processpointeritem|, which is responsible for data formating. 
    
    Macro |\showpointerlist| should be used without redefinition in the place
    where you want to display the content of its argument; for custom data
    formating redefine |\processpointeritem| macro.
         
  * |\processpointeritem| has three arguments (page-name, page-number, 
    destination) and visualizes these data. The arguments are delimited as 
    follows
    
    {\tt|\def\processpointeritem#1|\char124|#2|\char124|#3\endarg{...}|}

    An example definition may be

    {\tt|\def\processpointeritem#1|\char124|#2|\char124}|#3\endarg{%
  \hbox{\pdfstartlink attr {/Border [0 0 0]}% 
        goto name {#3} #2 (#1)\pdfendlink}%
}|

   (note how to use the {\it destination} argument), 
   or much simpler (if we don't need hyperlink features):

    {\tt|\def\processpointeritem#1|\char124|#2|\char124}|#3\endarg{%
  \hbox{#2 (#1)}%
}|
\endlist

For font management there are macros 

\list
* |\size[#1]| for size changes, and

* |\rm|, |\it|, |\bf|, |\ss|, |\si| for type face switching.
\endlist

If you use own texts in \TeX\ macros (like a label for the scale bar), these 
font switches work only in restricted manner: it's supposed that all characters 
are present in the first encoding specified in the initialization file. 

Now we're ready to define the |\dopage| macro. You may choose which of the 
predefined elements to use. A very simple example would be

|layout my_layout
  scale 1 200
  page-setup 29.7 21 27.7 19 1 1 cm
  size 26.7 18 cm
  overlap 0.5 cm

  \def\dopage{\box\mapbox}

  \insertmaps 

endlayout|

which defines the landscape A4 layout without the navigator nor any texts. There 
is only a map on the page.

Note the |\insertmaps| macro. Map pages are inserted at its position. [Similar 
macro |\insertlegend| will insert the map legend when Therion supports it.]
This is not done automatically because you may wish to insert some other pages 
before the first map page. (In this case something like |\input myfile.tex| 
should preceed the |\insertmaps| macro.) Without the |\insertmaps| invocation 
no map pages will be included!

More advanced is the default definition of the |\dopage| macro:

|\def\dopage{%
  \vbox{\line{\hfil\box\mapbox\hfil}
    \bigskip
    \line{%
     \vbox to \ht\navbox{
      \hbox{\size[20]\the\pagelabel
        \ifpagenumbering\space(\the\pagenum)\fi
        \space\size[16]\the\pagename}
      \ifpagenumbering
        \medskip
        \hbox{\qquad\qquad
          \vtop{%
  	    \hbox to 0pt{\hss\showpointer\pointerN\hss}
            \hbox to 0pt{\llap{\showpointer\pointerW\hskip0.7em}%
	           \raise1pt\hbox to 0pt{\hss$\updownarrow$\hss}%
	           \raise1pt\hbox to 0pt{\hss$\leftrightarrow$\hss}%
		   \rlap{\hskip0.7em\showpointer\pointerE}}
  	    \hbox to 0pt{\hss\showpointer\pointerS\hss}
	  }\qquad\qquad
          \vtop{
	    \def\arr{$\uparrow$}
	    \showpointerlist\pointerU
	    \def\arr{$\downarrow$}
	    \showpointerlist\pointerD
	  }
        }
      \fi
      \vss
      \scalebar{10}m
      }\hss
      \box\navbox
    }
  }
}|

Using other plain \TeX\ macros or \TeX\ primitives it's possible to add other 
features, e.g.~a different layout for odd and even pages; headers and footers;
or adding a logo to each page. 

\subchapter Customizing text labels.

There is a preliminary interface to changing font sizes for labels via the \MP\ 
macro

|fonts_setup(<tinysize>,<smallsize>,<normalsize>,<largesize>,<hugesize>);|

which may be used inside of the |code metapost| section of the |layout| 
command. |<normalsize>| applies to point label, |<smallsize>| applies to 
remark and all other point labels. Each of them may apply to line label 
according to its |-size| option.

Example:

|code metapost
  fonts_setup(6,8,10,14,20);|
  

\subchapter New map symbols.

Therion's layout command makes it easy to switch among various predefined map 
symbol sets. If there is no such symbol or symbol set you want, it's possible 
to design new map sybols. 

Howewer, this requires knowledge of the \MP\ language, which is used for map 
visualization. It's described in 

\list
  Hobby, J. D.: {\it A User's Manual for MetaPost}, available at
     \path|http://cm.bell-labs.com/cm/cs/cstr/162.ps.gz|
\endlist

User may also benefit from compehensive reference to the {\manfnt METAFONT} 
language, which is quite similar to \MP:

\list
  Knuth, D. E.: {\it The METAFONTbook}, Reading, Massachusetts, Addison-Wesley
    $^1$1986
\endlist

New symbols may be defined in the |code metapost| section of the |layout| 
command. This makes it easy to add new symbols at the run-time. It's also 
possible to add symbols permanently by compiling into Therion executable 
(see the {\it Appendix} for instructions how to do this).

Each symbol has to have a unique name, which consists of following items:

\list 
* one of the letters `p', `l', `a' for point, line or area symols, 
  respectively;
* underscore character;
* symbol type as listed in the chapter {\it Data format\/} with all dashes
  removed;
* if the symbol has a subtype, add underscore character and subtype;
* underscore character;
* symbol set identifier in uppercase
\endlist

Example: standard name for a point `water-flow' symbol with a `permanent' 
subtype in the `MY' set is |p_waterflow_permanent_MY|. 

Each new symbol has to be registered by a macro call

|initsymbol("<standard-name>");|

unless it's compiled into Therion executable. 

There are four predefined pens {\it PenA} (thickest) \dots\ {\it PenD} 
(thinest), which should be used for all drawings. 
For drawing and filling use |thdraw| and |thfill| commands instead of \MP's 
|draw| and |fill|.

\subsubchapter Point symbols.

Point symbols are defined as macros using |def ... enddef;| commands.
Majority of point symbol definitions has four arguments: 
position (pair), rotation (numeric), scale (numeric) and alignment (pair). 
Exceptions are {\it section} which has no visual represantation; 
all {\it labels}, which require special treatment as described in the
previous chapter, and 
{\it station} which takes only one argument: position (pair).

All point symbols are drawn in local coordinates with the length unit {\it u}. 
Recommended ranges are $\left<-0.5u,0.5u\right>$ in both axes. The symbol 
should be centered at the coordinates' origin.
For the final map, all drawings are transformed as specified in the {\it T\/} 
transformation variable, so it's necessary to set this variable before drawing. 

This is usually done in two steps (assume that four arguments are 
{\it P}, {\it R}, {\it S}, {\it A}): 
  
\list
* set the {\it U} pair variable to $\left({width\over2},{height\over2}\right)$ 
  of the symbol for correct alignment. The alignment argument {\it A} is a pair 
  representing ratios $\left(shift_x\over U_x\right)$ and
  $\left(shift_y\over U_y\right)$. 
  
  (Hence |aligned A| means |shifted (xpart A * xpart U, ypart A * ypart U)|.)
* set the {\it T} transformation variable 

  {\catcode`\=12|T:=identity aligned A rotated R scaled S shifted P;|}
\endlist

For drawing and filling use |thdraw| and |thfill| commands instead of \MP's 
|draw| and |fill|. These take automatically care of {\it T} transformation.

An example definition may be

|def p_entrance_UIS (expr P,R,S,A)=
  U:=(.2u,.5u);
  T:=identity aligned A rotated R scaled S shifted P;
  thfill (-.2u,-.5u)--(0,.5u)--(.2u,-.5u)--cycle;
enddef;
initsymbol("p_entrance_UIS");|

\subsubchapter Line symbols.

Line symbols differ from point symbols in respect that there is no local 
coordinate system. Each line symbol gets the {\it path} in absolute coordinates 
as the first argument. Therefore it's necessary to set {\it T} variable to 
|identity| before drawing.

Following symbols take additional arguments: 
\list
* arrow = numeric: 0 is no arrows, 1 arrow at the end, 2 begin, 3 both ends
* contour = text: list of points which get the tick or $-1$ to mark tick in the
    middle
* slope = numeric: 0 no border, 1 border; text: list of (point,direction,length) 
  triplets
\endlist

Usage example:

|def l_wall_bedrock_UIS (expr P) = 
  T:=identity;
  pickup PenA;
  thdraw P;
enddef;
initsymbol("l_wall_bedrock_UIS");|

\subsubchapter Area symbols.

Areas are defined as patterns with the same user interface (without the 
|patterncolor| macro) as described in 
the article

\list
  Bolek, P.: ``\MP\ and patterns,'' {\it TUGboat}, 3, XIX (1998), pp.~276--283,
  available online at
  \path|http://www.tug.org/TUGboat/Articles/tb19-3/tb60bolek.pdf|
\endlist

You may use standard \MP\ |draw| and similar macros without setting of {\it T} 
variable.

Example:

|beginpattern(a_water_UIS);
  draw origin--10up withpen pensquare scaled (0.02u);
  patternxstep(.18u);
  patterntransform(identity rotated 45);
endpattern;
initsymbol("a_water_UIS");|

\endinput
